---
title: "Week 9"
author: "Penelope Pooler Eisenbies"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=TRUE}

# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, 
                      highlight=T)

# suppress scientific notation
options(scipen=100,
        getSymbols.warning4.0 = FALSE)

# install helper package (pacman), if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
# pacman should be first package in parentheses and then list others
pacman::p_load(pacman,tidyverse, ggthemes, magrittr, knitr, lubridate, 
               tidyquant, highcharter, dygraphs, gridExtra, RColorBrewer, 
               flexdashboard)

# verify packages (comment out in finished documents)
p_loaded()

```


---

### Project Meetings

Thursday I will meet with each group after reading and providing feedback on proposal rough draft.

- Order will be determined **MOSTLY** randomly.

  - Schedule will be posted tomorrow.
  
  - If your class schedule or commitments mean that you can not be here exactly at 9:30 or you have to leave class early, please let me know **TODAY**.
  
<br>

- Come prepared to answer questions about your proposal and data sources.
  
- Also come prepared with questions for me.
  
- I will provide suggestions about 
  
  - How to focus or expand your plans
    
  - Additional data sources for your project 
    
  - Suggestions for dashboard visualizations or summaries
  that are good fit for your plan
  
- If needed, I will follow up with links to additional resources that will be helpful.
    
---
    
### Data Mgmt for HW 5 - Part 1 Dashboard 

#### New Skills introduced last week

- Using `ifelse` to create a categorical variable

  - Can be combined with `factor` to create a factor variable
  
  - Factor variables are useful for plots, tables, and analyses
  
    - Character (text variable) are easier to work with when managing and manipulating data.
    - Factor variables are preferred when plotting or presenting data in tables.
  
- Modifying and cleaning text variables

  - `tolower` and `toupper`
  
  - `str_trim` (as well as `str_squish` and `str_pad`)
  
  - `gsub` and `ifelse`
  
---

### Data Mgmt for HW 5 - Part 1 Dashboard 

#### Review of Previous Data Mgmt skills used

- `filter` and `select`
  
- `group_by` and `summarize`
  
- `pivot_wider` and `pivot_longer`

<br>

**Note:** Today we will cover how to use `pivot_wider` and `pivot_longer` to ***complete*** and incomplete dataset.

In HW 5 - Part 1:

- I provide R code in the dashboard example to complete incomplete data

- I also provide code to complete this task using `pivot_wider` and `pivot_longer` in an Optional R Script.

- Not needed for Netflix Movie data even though these data are not fully complete.
  
---
  
### Skills Covered in HW 5 - Part 1


- **Panel 1 - Interactive Time Series Plots** 

   - Review of importing and plotting xts datasets
  
   - `xts` stands for extensible time series
  
   - Multiple `xts` plot objects can be displayed using `hw_grid`
  
- You should also review (Not in Panel 1):

   - Converting a `tibble` to an `xts` 
  
   - Converting an `xts` to a `tibble`
  
- New Skills

   - [Using links to cite data sources](https://github.com/rstudio/cheatsheets/raw/main/rmarkdown.pdf)


---

### Skills Covered in HW 5 - Part 1
  
- **Panel 2 - Grouped Stacked Barplots with Facets and Summary Table**

  - Review of Previously Covered Skills
  
     - `pivot_longer`to reshape `nflx_mv` data
    
     - `ifelse` is used to simplify `release period` data

     - `mutate` and `factor` is used to create two factor variables with specified levels
    
     - `filter` is used to subset the data to release periods with enough data.
  
  - New Skills - Panel 2 Plot is a grouped stack bar plot that uses a new option, `facet_grid`
    
     - `facet_grid(~release_period)` specifies side-by-side plots, 1 for each release period.
    
     - `scale_fill_brewer(palette = "Spectral")` allows the user to specify a color palette
    
     - Theme and palette options (Optional Extra Credit)
    
        - [ggtheme options](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)
      
        - [Palette options](http://www.sthda.com/english/wiki/colors-in-r#use-wes-anderson-color-palettes) and another link with a [comprehensive list of palettes](https://github.com/EmilHvitfeldt/r-color-palettes)
      
---

### Skills Covered in HW 5 - Part 1

- **Panel 2 - More about the Bar Plot and Summary Table**
    
   - Be aware of how the barplot is specified in the `geom_barplot ` statement.
  
   - Stacked bars were used to save space.
    
- Panel 2 Table
  
   - The skills used to create to select and summarize data are review.
    
   - The table is presented using `kable` (also review).
    
   - Note that for projects, you can refine your table further, if needed, using commands from `kableExtra` package.


---

### Skills Covered in HW 5 - Part 1

- **Panel 3 - Area Plot**

  - Useful new Skill:
  
     - summing across rows: `mutate(total = sum(c_across(comedies:kids)))`
    
     - sometimes used with `rowwise()` and `ungroup` but that is not needed here
    
  - Area Plot (Review)
  
     - Similar to line plot but uses `geom_area(aes(x = , y = , fill=))`
    
     - `scale_x_continuous(breaks = )` used to show all years in data
    
     - `scale_fill_brewer` used to specify same palette as Panel 2

---

### Week 9 In-class Exercises 

- This short exercise demonstrates why a **complete** dataset with all combination levels is needed for area plots.

- These steps are similar to the Optional R Script for HW 5 - Part 1 

**`1.`** Import gt_cars dataset and examine the data

```{r import raw data and examine them}
my_gtcars <- read_csv("gt_cars.csv", show_col_types = F) |> glimpse()

```


---

### Week 9 In-class Exercises 

**`2.`** Select the following variables

   - year, bdy_style (body style), drivetrain (drive train)

**`3.`** Group data by year, body style, and drive train and count number of observations

   - command to summarize number of observations is `n()`

```{r select vars then group data and summarize}
my_gtcars <- read_csv("gt_cars.csv", show_col_types = F) |>
  select(year, bdy_style, drivetrain) |>
  group_by(year, bdy_style, drivetrain) |>
  summarize(n = n())                              # new command n()
head(my_gtcars)                                   # incomplete dataset

```

---

### Week 9 In-class Exercises 

**`4.`** Use pivot_wider to create a column for each drivetrain

**`5.`** Replace missing values with 0

```{r reshape and replace NAs with 0s}
my_gtcars <- my_gtcars |> 
  pivot_wider(id_cols= c("year", "bdy_style"), names_from = drivetrain, values_from = n)
my_gtcars[is.na(my_gtcars)] <- 0  # convert missing values to 0
head(my_gtcars)

```

---

### Week 9 In-class Exercises (L15 - Q1)

***PointSolutions Session ID: bua455s22***

**`6.`** Sum across rows to total data from both drive trains

- Q1: What is the command to sum across rows?
   
  - `drivetrain` will be dropped in next step; 
  
  - included here to demonstrate sum across
   
```{r sum accross rows to create total column}
my_gtcars <- my_gtcars |>
   mutate(total = sum(c_across(rwd:awd))) 

```



---

### Week 9 In-class Exercises (L15 - Q2 & Q3)

***PointSolutions Session ID: bua455s22***

**`7.`** Select year, body style and total

- Q2: How many observations do you have?
   
- Q3: If every body style was made in every year, how many observations would you have?

  - There are 4 body styles and 4 years represented in these data
  
- **Also in this chunk:** Command to capitilize first letter of words in a string: `str_to_title`
   
```{r select vars of interest and examine}

my_gtcars <- my_gtcars |> select(year, bdy_style, total) |>
  mutate(bdy_style = str_to_title(bdy_style)) |>    # Capitalize first letters
  glimpse()

```


---

**`8.`** Fill in dataset by using pivot_wider and then pivot longer

Complete code is relatively short and efficient.

- Factors should be created **AFTER** this process, if needed.

```{r complete conversion code}

my_gtcars1 <- my_gtcars |>
  
   pivot_wider(id_cols= year, 
               names_from = bdy_style, 
               values_from = total) |>
  
   pivot_longer(cols=Coupe:Sedan, 
                names_to = "bdy_style", 
                values_to = "total")

my_gtcars1[is.na(my_gtcars1)] <- 0
my_gtcars1                              # show final dataset

```

---

### Comparison of Complete and Incomplete Data Area Plots

```{r area plot code}

(gt_car_area_plot <- my_gtcars |>
  ggplot() +
  geom_area(aes(x=year, y=total, fill=bdy_style)) +
  theme_classic() +
  theme(legend.position="bottom") +
  scale_x_continuous(breaks=seq(2014,2017,1)) +
  scale_fill_brewer(palette = "Spectral") +
  labs(x="Year", y="Number of cars", fill="",
       title="Number of Cars of Each Body Type",
       subtitle="Incomplete Data for 2014 - 2017",
       caption="Data Source: https://CRAN.R-project.org/package=gt"))

(gt_car_area_plot1 <- my_gtcars1 |>
  ggplot() +
  geom_area(aes(x=year, y=total, fill=bdy_style)) +
  theme_classic() +
  theme(legend.position="bottom") +
  scale_x_continuous(breaks=seq(2014,2017,1)) +
  scale_fill_brewer(palette = "Spectral") +
  labs(x="Year", y="Number of cars", fill="",
       title="Number of Cars of Each Body Type",
       subtitle="Complete Data for 2014 - 2017",
       caption="Data Source: https://CRAN.R-project.org/package=gt"))

```

---

### **Key Points from Week 9**


**Data Management Skills - Some Review and Some New**

- Some Key New Data Management Skills:

  - summarizing number of observations by category using `n()`
  
  - Using `sum(c_across())` to sum across rows
  
  - Knowing how to ***complete*** data that is incomplete 

- HW 5 - Part 1 is due Wedneday, 11/2.

- Next week 

  - Practice questions for Quiz 2 (Quiz is on Thu. 11/10)
  
  - Managing and presenting geographic data in choropleth maps.


**You may submit an 'Engagement Question or Comment' about Week 9 lecture until Thursday, 10/25, at midnight on Blackboard.**